---
title: "Creating your own functions in R"
author:
  - "Luke W. Johnston"
date: "2019-11-29"
categories:
  - Beginner-Intermediate
  - R
  - Code-along
tags:
  - functions
  - productivity
  - efficiency
  - tidy code
  - don't repeat yourself
---

> NOTE: This is still in development.

```{r setup, echo=FALSE}
source(here::here("R/utils-for-content.R"))
knitr::opts_chunk$set(eval = FALSE)
```

## Resources for learning and help

[Here for resources](/more-info/#resources)


## Session content

> This is the code used during the session. I've added some comments and more 
explanation to the code.

### All actions in R are functions

The `+` is a function, `mean()` is a function, `[]` is a function... everything 
that does something is called a function in R. So this to add 1 with 1:

```{r}
1 + 1
```

... is a function that takes 1 and adds 1 to it. Functions have the basic 
structure of:

- the name of the new function `add_nums <-`
- the function to create the new function `function()`, along with the arguments
`num1`, `num2`
- the code to do the action of the function, everything between `{}`
    - should have a `return()` at the bottom that says what the function creates

```{r}
add_nums <- function(num1, num2) {
    stopifnot(is.numeric(num1), is.numeric(num2))
    added <- num1 + num2
    return(added)
}
```

You can use the new function by running the above code and writing out your new 
function, with arguments to give it.

```{r}
add_nums(1, 2)
add_nums(c(1:10), 2)
# Two numbers should be the same length vector (same number of numbers) or one
# should be only one number.
add_nums(c(1:10), c(1, 4, 6))
add_nums(c(1:10), c(11:20))
```

There are a few things to consider. In R there are different "methods" of functions.
This is way above what is necessary for this session, but if you are curious
[this website](http://adv-r.had.co.nz/S3.html) has a great explanation of the
different methods (e.g. S3 methods). Be warned, the website is fairly advanced!

You can always look at the contents of all functions in R. So an example of an
S3 function:

```{r}
# Generic S3
print
# Printing for data.frames
print.data.frame
```

Ok, let's get to something a bit more interesting. Usually we create plots that
are more or less the same each time, but with different variables or data.
So this is a great example of using a function to simplify your code. Let's load
up the ggplot2 package for plotting and the
[gapminder](https://github.com/jennybc/gapminder) dataset.

```{r}
library(ggplot2)
library(gapminder)
head(gapminder)
```

Let's plot year by life expectancy:

```{r}
ggplot(gapminder, aes(x = year, y = lifeExp)) +
    geom_smooth()
```

What if we wanted to see another plot by pop over time:

```{r}
ggplot(gapminder, aes(x = year, y = pop)) +
    geom_smooth()
```

Or another plot... and so on. This starts getting a bit tedious, as you are just
copying and pasting. There is a better way! Convert it into a function! The
typical process for converting code into a function is first to write the code
and make sure it works. Then wrap it in a function. And start replacing the
variable names with the arguments. So:

```{r, eval=FALSE}
# First this:
ggplot(gapminder, aes(x = year, y = pop)) +
    geom_smooth()

# Then this:
plot_smooth <- function() {
    plot <- ggplot(gapminder, aes(x = year, y = pop)) +
        geom_smooth()
    return(plot)
}

# Then lastly this:
plot_smooth <- function(data, xvar, yvar) {
    plot <- ggplot(data, aes(x = xvar, y = yvar)) +
        geom_smooth()
    return(plot)
}
```

But, this function won't work! That's because there is a tricky bit that you
will quickly encounter in R... And that is called non-standard evaluation (NSE; check
out [here](https://cran.r-project.org/web/packages/lazyeval/vignettes/lazyeval.html)
or [here](https://dplyr.tidyverse.org/articles/programming.html) for more indepth
look at what non-standard evaluation is). Because ggplot2 uses NSE, you will
have to do things slightly differently. The `aes` in ggplot2 uses NSE. So you have 
to use `aes_string` instead.

```{r}
plot_smooth <- function(data, xvar, yvar) {
    plot <- ggplot(data, aes_string(x = xvar, y = yvar)) +
        geom_smooth()
    return(plot)
}
plot_smooth(gapminder, "year", "lifeExp")
plot_smooth(gapminder, "year", "pop")
```

If you want to make sure that who ever uses your function will not use a wrong 
argument, you can use "defensive programming" via the `stopifnot()` function.
This forces the code to only work if `xvar` and `yvar` are character (e.g. `"this"`)
argument.

```{r}
plot_smooth <- function(data, xvar, yvar) {
    stopifnot(is.character(xvar), is.character(yvar))
    plot <- ggplot(data, aes_string(x = xvar, y = yvar)) +
        geom_smooth()
    return(plot)
}
```

This NSE evaluation also happens in a popular package called dplyr. If we wanted
to do a common analysis or data wrangling like so:

```{r}
library(dplyr)
gapminder %>% 
    select(continent, year, pop) %>% 
    group_by(continent, year) %>% 
    summarise(mean(pop))
```

... and do this again but change `pop` to another variable, we can create a function.

```{r}
# First:
gapminder %>% 
    select(continent, year, pop) %>% 
    group_by(continent, year) %>% 
    summarise(mean(pop))

# Then:
mean_by_continent <- function() {
    by_continent <- gapminder %>%
        select(continent, year, pop) %>%
        group_by(continent, year) %>%
        summarise(mean(pop))
    return(by_continent)
}

# Finally:
mean_by_continent <- function(data, variable) {
    by_continent <- data %>%
        select(continent, year, variable) %>%
        group_by(continent, year) %>%
        summarise(mean(variable))
    return(by_continent)
}
```

... then use dplyr's standard evaluation functions, which are usually the function
with a `_at` or `_if` or other variation of that at the end...

```{r}
mean_by_continent <- function(data, variable) {
    by_continent <- data %>%
        select_at(c("continent", "year", variable)) %>%
        group_by(continent, year) %>%
        # This needs to change order a bit, since the mean will be applied to
        # the variable.
        summarise_at(variable, mean)
    return(by_continent)
}
```

And we can use it:

```{r}
mean_by_continent(gapminder, "lifeExp")
mean_by_continent(gapminder, "pop")
mean_by_continent(gapminder, c("pop", "lifeExp"))
```

Send the output to create a table!

```{r}
mean_by_continent(gapminder, "lifeExp") %>% 
    head() %>% 
    knitr::kable(caption = "Table caption here")
```

We can even use both our functions together!

```{r}
mean_by_continent(gapminder, "lifeExp") %>% 
    plot_smooth("year", "lifeExp")
```

