---
title: "Creating your own functions in R"
author:
  - "Luke W. Johnston"
date: "2019-11-29"
categories:
  - Beginner-Intermediate
  - R
  - Code-along
tags:
  - functions
  - productivity
  - efficiency
  - tidy code
  - don't repeat yourself
description: >
  Do you write code multiple times that only slightly changes each time you write
  (or copy and paste) it? Learn how to not repeat yourself by writing your own
  functions!
---

```{r setup, include=FALSE}
source(here::here("R/utils-for-content.R"))
```

## Resources for learning and help

[Here for resources](/more-info/#resources)

## Session content

Before we begin, we'll set up an R project, create a folder named `R/`, and then
create a new file in the `R/` folder of the project.

In R, all actions are functions. For instance, the `+` is a function, `mean()`
is a function, `[]` is a function, and so on. So creating your functions can
make doing your work easier and more efficient. Making functions always has a
basic structure of:

1. Giving a name to the function (e.g. `mean`).
2. Starting the function call using `function()`, assigning it to the name with
`<-`. This tells R that the name is a function object.
3. Optionally providing arguments to give to the function call, for instance
`function(arg1, arg2, arg3)`. 
4. Filling out the body of the function, with the arguments (if any) contained
inside, that does some action.
5. Optionally, but strongly encouraged, use `return` to indicate what you want 
the function to output.

While there is no minimum or maximum number of arguments you can provide for a
function (e.g. you could have zero or dozens of arguments), its good practice
for yourself and for others if you have as few arguments as possible to get
the job done.

So, the structure is:

```{r function-structure, eval=FALSE}
name <- function(arg1, arg2) {
  # body of function
  ... code ....
  return(output)
}
```

... and an example:

```{r create-add-function}
add_nums <- function(num1, num2) {
  added <- num1 + num2
  return(added)
}
```

You can use the new function by running the above code and writing out your new 
function, with arguments to give it.

```{r call-function}
add_nums(1, 2)
```

The function name is fairly good... `add_nums` can be read as "add numbers".
It's also good practice to add some formal documentation to the function. Use
the "Insert Roxygen Skeleton" in the "Code" menu list of RStudio (or by typing
`Ctrl-Shift-Alt-R`) and you can add template documentation right above the
function. It looks like:

```{r roxygen-docs-1}
#' Title
#'
#' @param num1 
#' @param num2 
#'
#' @return
#' @export
#'
#' @examples
add_nums <- function(num1, num2) {
  added <- num1 + num2
  return(added)
}
```

In the `Title` area, this is where you type out a brief sentence or several words 
that describe the function. Creating a new paragraph below this line allows you 
to add a more detailed description. The other items are:

- `@param num` lines are to describe what each argument is for.
- `@return` describes what output the function provides. Is it a data.frame? A plot? 
What else does the output give?
- `@export` tells R that this function should be accessible to the user of your
package. Delete it for non-package functions.
- `@examples` lines below this are used to show examples of how to use the function.
Very useful when making packages, but not really in this case.

```{r roxygen-docs-2}
#' Add two numbers together.
#'
#' This is just an example function to show how to create one.
#'
#' @param num1 A number here.
#' @param num2 A number here.
#'
#' @return Returns the sum of the two numbers.
#'
add_nums <- function(num1, num2) {
  added <- num1 + num2
  return(added)
}
```

This is the very basics of making a function. But let's get to something more
useful!

### Functions to wrangle

Let's do something you might actually do in a data analysis project. Let's 
suppose you want to create a scatter plot of two variables in a dataset, that you may
want to include as a figure in a report or maybe to use while you explore the
data. 

A typical workflow for creating a function is to:

1. Write the code as you normally would and make sure it does what you want.
1. Set up the function structure and put the code into the function. Make sure
the function name explains what the function does.
1. Create the arguments you want and replace the variables in the code with the
arguments. Make sure the argument names explain what they are used for.
1. Append `packagename::` to any function inside your function that comes from a
package (e.g. if the function comes from ggplot2, it will look like `ggplot2::geom_point()`).
This ensures that R doesn't mistake the function for one in another package.
1. Add a return command to the end.
1. Run the function and test it out to see if it works. If not, try to debug the
code by adding `print()` commands inside the function to see what is happening.
1. Once it all works as you want it, try the function using different arguments.
If it doesn't work, try 5 again.
1. Finally, add the documentation to the function so you remember what it was for.
1. To make your function more "secure", you could be "defensive" and add checks 
that your arguments are used appropriately.

Alright, let's try with making a graph. We'll use the NHANES dataset from the NHANES
package, plus we're going to use ggplot2, so let's load them up!

```{r load-packages}
library(NHANES)
library(ggplot2)
```

Then we create a basic scatter plot:

```{r scatter-plot}
ggplot(NHANES, aes(x = Height, y = Weight)) +
  geom_point()
```

If we wanted to make another scatter plot of some of variables, we would normally
have to type out all that code again. But instead, we could make a function of
it! So let's get it into function form:

```{r plot-function-template, eval=FALSE}
scatter_plot <- function() {
  scatter <- ggplot(NHANES, aes(x = Height, y = Weight) +
    geom_point()
  return(scatter)
}
```

Then let's add in the arguments and try it out!

```{r plot-function-arguments}
scatter_plot <- function(.dataset, .xvar, .yvar) {
  scatter <- ggplot(.dataset, aes(x = .xvar, y = .yvar)) +
    geom_point()
  return(scatter)
}
```

I created arguments that begin with `.`. This is mostly personal preference, but
I like it because it makes it easier to see in the code what is an argument for
this function. But you could just as easily not use the `.`. Ok, let's try out the
function.

```{r try-out-function, error=TRUE}
scatter_plot(NHANES, Height, Weight)
```

Oops! What happened? We had an error... We've actually encountered a problem from
something called "[non-standard evaluation]" (or NSE). NSE is very commonly used
in most tidyverse packages like ggplot2, as well as throughout base R. When
creating functions, NSE can be difficult to work around, but for everyday use 
it is incredibly useful! 

[non-standard evaluation]: http://adv-r.had.co.nz/Computing-on-the-language.html

For instance, NSE is what allows you to use formulas
(e.g. `y ~ x + x2` in modelling) or allows you to type out `library(ggplot2)`.
In other programming languages, it would be `library("ggplot2")` and the use of
formulas wouldn't work. So it gives a lot of flexibility when doing data
analysis, but can give some headaches when programming in R. 

The reason it doesn't work here is R thinks that it needs to look for the `Height`
variable as an object in R... but what we *mean* is to use the `Height` as a
variable *in the dataset*. This applies for `Weight` as well. In normal ggplot2,
it uses NSE to see `Height` and interpret it was us meaning *Height in the dataset*.

To work around this, thankfully the developers of tidyverse created a way to
easily work with NSE in functions, with the "curly-curly" (`{{ }}`) function. We
fix the problem by wrapping the arguments in the function with curly-curly:

```{r plot-function-curly-curly}
scatter_plot <- function(.dataset, .xvar, .yvar) {
  scatter <- ggplot(.dataset, aes(x = {{.xvar}}, y = {{.yvar}})) +
    geom_point()
  return(scatter)
}
scatter_plot(NHANES, Height, Weight)

# Try with different variables
scatter_plot(NHANES, Age, Height)
```

Nice! Ok, next step is to be explicit about using ggplot2 for functions. This is
a good practice to use after you've created your function. Explicitly
indicating which functions come from which package is important, since you
shouldn't use `library()` calls in your function. We be explicit by using the
`packagename::function_name()` format (you don't need to do this for base R
functions, only those from packages). 

```{r plot-function-ggplot2-explicit}
scatter_plot <- function(.dataset, .xvar, .yvar) {
  scatter <- ggplot2::ggplot(.dataset, ggplot2::aes(x = {{.xvar}}, y = {{.yvar}})) +
    ggplot2::geom_point()
  return(scatter)
}
# Check that it still works.
scatter_plot(NHANES, Height, Weight)
```

Then, let's add the roxygen documentation (`Ctrl-Alt-Shift-R`) and fill it in:

```{r plot-function-documentation}
#' Create a scatter plot of two continuous variables.
#'
#' @param .dataset Data to plot.
#' @param .xvar Variable for the x axis.
#' @param .yvar Variable for the y axis.
#'
#' @return Creates a ggplot.
#'
scatter_plot <- function(.dataset, .xvar, .yvar) {
  scatter <- ggplot2::ggplot(.dataset, ggplot2::aes(x = {{.xvar}}, y = {{.yvar}})) +
    ggplot2::geom_point()
  return(scatter)
}
```

If you want to make sure that who ever uses your function will not use a wrong 
argument, you can use "defensive programming" via the `stopifnot()` function.
For instance, for this function you can say, always include the x and y variables
(don't have them missing):

```{r plot-function-stopifnot}
#' Create a scatter plot of two continuous variables.
#'
#' @param .dataset Data to plot.
#' @param .xvar Variable for the x axis.
#' @param .yvar Variable for the y axis.
#'
#' @return Creates a ggplot.
#'
scatter_plot <- function(.dataset, .xvar, .yvar) {
  stopifnot(!missing(.xvar), !missing(.yvar))
  scatter <- ggplot2::ggplot(.dataset, ggplot2::aes(x = {{.xvar}}, y = {{.yvar}})) +
    ggplot2::geom_point()
  return(scatter)
}
scatter_plot(NHANES, Height, Weight)
```

There are many more tools for "defensive programming" and it could be a session
on its own, but here we just introduce it.

A massive advantage of using functions is that if you want to make a change to all
your code, you can very easily do it in the R function and it will change all 
your other code too! So let's say we want to change the plot theme, so it looks
better:

```{r plot-function-add-more}
#' Create a scatter plot of two continuous variables.
#'
#' @param .dataset Data to plot.
#' @param .xvar Variable for the x axis.
#' @param .yvar Variable for the y axis.
#'
#' @return Creates a ggplot.
#'
scatter_plot <- function(.dataset, .xvar, .yvar) {
  stopifnot(!missing(.xvar), !missing(.yvar))
  scatter <- ggplot2::ggplot(.dataset, ggplot2::aes(x = {{.xvar}}, y = {{.yvar}})) +
    ggplot2::geom_point() +
    ggplot2::theme_minimal()
  return(scatter)
}
scatter_plot(NHANES, Height, Weight)
```

You now have an idea on how to make a basic function!
