---
title: "2020-01-19-intermediate-dplyr"
author: 
- "Sara Tomiolo"
- "Luke Johnston"

date: "1/17/2020"
description: >
  Learn to reshape data in long or wide form, manage factors and dates using the packages dplyr, tidyr, lubridate, forcats and stringr.
categories:
  - Intermediate
  - R
  - Code-along
tags:
  - tidyr
  - dplyr
  - forcats
  - stringr
  - lubridate
  - data manipulation
  - factors
  - dates
---


```{r setup, echo=FALSE}
source(here::here("R/utils-for-content.R"))
```

## Preparation

Please install the tidyverse package:

```{r, eval=FALSE}
install.packages("tidyverse")
library(lubridate)
```

### Prerequisites

In this lesson we assume you are familiar with the basic functions of dplyr (especially group-by() and summarize()), if not we recommend you to take a look at the [intro_dplyr lesson](https://au-cru.github.io/site/material/2019-11-15-intro-dplyr/) 


### Objectives

I know the cognitive load is insane, I am just jotting down ideas and the how can I make them work together, then I will select 3 objectives out of all this and cut the rest. 
1. To learn to restructure your data between long and wide format with pivot_longer() and pivot_wider() functions from tidyr package.
2. Learn to combine group_by() and summarize() function with pivot_longer() and pivot_wider() functions.
3. Handle factor variables using functions.... from the forcacts package.
4. Manage dates with lubridate functions.
5. Create new variables by joining or separating previously existing variables using the functions of the package stringr.
    
**At the end of this session you will be able:**

1. To restructure your data in long or wide format.
2. To reshape data by using grouping, summarizing and reshaping dataframes.
2. To manage and reorder the level of factorial variables.
3. Reformat dates with lubridate

### Why learning to reformat and reshape data?

Often we receive databases that have been created by others, and in  whcih information is organized in a "messy" way. Some other time, we simply need to reorder the structure of the data to be able to carry out certain data analysis or data visualization. Tidyverse, provide some very powerful packages that can allow for complete reorganization of our data in a few lines of codes. Using these packages and their functions has several advantages:

- It makes our lives easier and our work faster, because we  can jump between different formats of our data in an agile and easy to track way;

- It makes our work enterely reproducible.


