---
title: "Data visualization with ggplot2, part 2"
author:
  - "Luke W. Johnston"
  - "Sara Tomiolo"
date: "2020-01-31"
description: >
  Dig deeper into making the most of making data visualizations with ggplot2.
  We'll show you how to plot multiple panels within a single plot, how to append
  multiple plots together, and how to make your plot more visually appealing.
categories:
  - Intermediate-Beginner
  - R
  - Code-along
tags:
  - ggplot2
  - data visualization
  - colour palettes
  - patchwork
---

```{r setup, echo=FALSE}
source(here::here("R/utils-for-content.R"))
```

# Session details

## Preparation

Please install these packages:

```{r pkg-install, eval=FALSE}
install.packages(ggplot2)
install.packages(patchwork)
install.packages(viridis)
```

## Objectives

By the end of this session you will know how to:

1. Apply facetting functions to include multiple "panels" on a single plot.
1. Consider the challenges with using colours and to make use of appropriate colour palettes.
1. Modify the theme settings to make visually appealling plots.
1. Discover and use new theme settings to modify nearly all aspects of a plot.

**Resources** for continuing learning and help are available on
the [More Info](/more-info/#resources) page.

# Making more use out of ggplot2

## First steps: Project setup and exploring the data

Like most teaching sessions, we'll start with creating an R project. 
R projects are incredibly useful tools and we strongly recommend you use them.
Create a new project, name it "more-ggplot2". 
Once the R project has been opened, create two new files and save them as 
"code-along.R" and "exercises.R".

Next, let's get familiar with the data we'll be using for this session, 
which is the `economics` dataset. 

```{r}
# Variables
names(economics)

# Content
head(economics)
```

### Exercise: Choose a dataset and check out the different classes of variables

There are several exercises in this session. Choose one of the below datasets
and use that dataset for all later exercises.

For complete R beginners, use:

- `mpg`

For more confident R users, use one of these:

- `economics`
- `diamonds`
- `msleep`
- `txhousing`

Check out the contents of the dataset you choose using.

Often the type of graph you decide to use, depends on the type of variable you want to represent. 

What type of variables do you see in your dataset? What is the meaning of different variable types? 

```{r, eval=FALSE}
# variable names of dataset
names(___)

# contents of dataset
str(___)

# summary of dataset
summary(___)
```


## Basic structure of using ggplot2

As a refresher, we'll cover the basic structure of ggplot2.
ggplot2 uses the "Grammar of Graphics" (gg). This is a powerful approach to
creating plots because it provides a consistent way of telling ggplot2 what to
do. There are at least three aspects to using ggplot2 that relate to the grammar:

- Aesthetics, `aes()`: How data should be mapped to the plot. Includes what to
put on x axis, on the y axis, colours, size, etc.
- Geometries, `geom_`: The visual representation of the data, as a layer. This
tells ggplot2 how to show the aesthetics. Includes points, lines, boxes, etc.
- Themes, `theme_` or `theme()`: How the plot should look like. Using this function we can customize several components of a graph, such as axis, lines, grids, plot background, position, size and font of the axis text and labels
- Title and labels can be added using the functions `ggtitle()` and `labs()` respectively

The aesthetics and the geometries are fundamental to create a graph, whereas scales, themes, and labels are feature that can be stacked in successive layers to improve the looks and the clarity of a graph.


**Tip:** To maximise the power of ggplot2, make heavy use of autocompletion. You can do 
this by typing, for instance, `geom_` and then hitting the TAB key to see a list
of all the geoms. Or after typing `theme()`, hit TAB to see all the options inside
theme.

## Best practices for creating graphs in R

It is relatively easy to create complex graphs in R, but all the features available can become quite distracting and lead to overly complicated and confusing graphs.

When creating graphs there are a few golden rules that should be kept in mind:
- When possible use raw data

- Avoid / remove unnecessary information

- Use colour in a meaningful way

<!-- Quickly go over RStudio IDE in relation to plotting -->
<!-- Get them to open two R scripts, name and save as "code-along" and "exercise" -->


## Axis titles and the theme

Let's get to making the plot prettier. First let's create a graph combining four different variables.

```{r}
co2_plot <-
    ggplot(CO2, aes(
        x = conc,
        y = uptake,
        #represents the combination of the two variables as different         #colours
        colour = paste(Treatment, Type)
        )
    ) +
    geom_point() +
    geom_smooth()


```


Then let's change its appearance. 
There are many many options to customise
the plot using the `theme()`. 
```{r}
# Some pre-defined themes
co2_plot + theme_bw()
co2_plot + theme_minimal()
```


Another feature of plot that can greatly improve the clarity of the 
graph is the colour palette
```{r, fig.width=7}
pretty_plot <- co2_plot +
    theme_classic() +
    scale_color_brewer(name = "Treatment and origin", palette = "Dark2") +
    # Find this information in ?CO2
    labs(x = "CO2 concentration (mL/L)",
         y = "CO2 update rate (umol/m2)") +
    theme(
        # all axis lines, must use element_line
        axis.line = element_line(colour = "grey50", size = 0.5),
        # all axis text, must use element_text
        axis.text = element_text(family = "sans"),
        # all axis tick marks, use element_blank to remove
        axis.ticks = element_blank()
    )
pretty_plot
```

### Exercise: Change theme items of the plot

Time: 10 min

```{r, eval=FALSE}
# use dataset with two continuous variables
ggplot(___, aes(x = ___, y = ___)) +
    # finish the geom to create either a point, smooth, or line layer
    ___ +
    # choose either a minimal, dark, light, or classic defined theme
    ___ +
    theme(
        # choose colours such as red, blue, black, grey, yellow, green
        # choose size from 2 to 8
        panel.grid.major = element_line(colour = ___, size = ___),
        # choose family such as sans, serif, Arial, Times New Romans
        axis.text = element_text(colour = ___, size = ___, family = ___)
    )
```




## Facetting: adding even more variables to a plot


You can also add another variable dimension by "facetting", which means splitting
the data up by the variable and plot by that variable.


```{r plot_facet}
colour_plot_mixed +
    geom_boxplot() +
    # Cols means to have them side by side (horizontal)
    # vars() is necessary to access variable from dataset
    facet_grid(cols = vars(SurveyYr))
# Or by another variable
colour_plot_mixed +
    geom_boxplot() +
    # Rows means to have them stacked vertically
    facet_grid(cols = vars(SurveyYr), rows = vars(PhysActive))
```

And you can add another geom as a layer on top of the previous one by adding
(`+`) a geom to the next line.

```{r plot_more_layers}
# Three layers
colour_plot_nums + 
    geom_point() +
    geom_smooth() +
    facet_grid(cols = vars(SurveyYr))
```

TODO: Add some code about using pivot_ to make use of facet

### Colours: Make your graph more accessible

Colour blindness is common in the population, and red-green colour blindness in
particular affects 8% of men and 0.5% of women. Making your graph more
accessible to people affected by colour blindness will also usually improve the
interpretability of your graphs for all people. For more detail on how colours
look to those with colour-blindness, see this
[documentation](https://cran.r-project.org/web/packages/viridis/vignettes/intro-to-viridis.html) 
from the [viridis] package. The viridis colour scheme and R package was
specifically designed to represent data well to all colour visions. There is
also a really good, informative [talk on
YouTube](https://www.youtube.com/watch?v=xAoljeRJ3lU) on this topic.

[viridis]: https://cran.r-project.org/packages=viridis

When using colours, take time to think about what you are trying to convey in
your figure and how your choice of colours will be interpreted. You can use
built-in colour schemes, or set your own. Let's stick to using builtin ones.
There are two, the viridis and the [ColorBrewer] colour schemes. Both are well
designed and are colour-blind friendly.

[ColorBrewer]: http://colorbrewer2.org/

```{r changing_color_schemes}
base_boxplot <- NHANES %>% 
    ggplot(aes(x = HomeOwn, fill = Education)) +
    geom_bar(position = position_dodge()) 

# Add the viridis scheme
base_boxplot +
    # _d() is for discrete. _c() is for continuous.
    scale_fill_viridis_d()
# Use another viridis color scheme
# Ranges from A to E
base_boxplot +
    scale_fill_viridis_d(option = "A")
# Or use the brewer scheme
base_boxplot +
    scale_fill_brewer()
base_boxplot +
    scale_fill_brewer(type = "qual")
```

See all colour palettes for the Brewer palette, use
`RColorBrewer::display.brewer.all()`. If you use a `colour = VariableName` in
`aes()`, you'll need to use `scale_colour_*` (either brewer or viridis).

### Exercise: Create a three (or more) variable plots

Time: 10 min

In the `exercises-visualizing.R` script, add a new Section (`Ctrl-Shift-R`) and
copy the code below into the script. Then, create (with your own data or with
NHANES) several plots with:

1. Two continuous (x, y) and one discrete variables (colour), with another
discrete variable as the facet. (Optionally include another layer, like
smoothing).
2. Three continuous variables (x, y, and colour), using the viridis colour
palette for continuous data.
3. Three discrete variables (x, fill, facet), using the brewer colour palette.

```{r 3d-more-exercise, eval=FALSE}
# 1. Two continuous, discrete, and one facet.
ggplot(___, aes(x = ___, y = ___, colour = ___)) +
    ___ +
    facet_
    # Optional: add another layer

# 2. Three continuous, with viridis
ggplot(___, aes(x = ___, y = ___, colour = ___)) +
    ___ +
    scale_

# 3. Three discrete, with brewer
ggplot(___, aes(x = ___, fill = ___)) +
    ___ +
    facet_ +
    scale_
```

<details><summary><strong>Click for a possible solution</strong></summary>
<p>

```{r 3d-more-answer, eval=FALSE}
# 1. Two continuous, discrete, and one facet.
ggplot(NHANES, aes(x = Height, y = Weight, colour = Gender)) +
    geom_point() +
    facet_grid(rows = vars(Diabetes))
    # Optional: add another layer

# 2. Three continuous, with viridis
ggplot(NHANES, aes(x = Height, y = Weight, colour = Age)) +
    geom_point() +
    scale_colour_viridis_c()

# 3. Three discrete, with brewer
ggplot(NHANES, aes(x = Gender, fill = Education)) +
    geom_bar(position = position_dodge()) +
    facet_grid(cols = vars(Diabetes)) +
    scale_fill_brewer()
```

</p>
</details>

## Themes: customize text, line, background and legends

There are so so so many options to modify the figure, and they are all basically
contained in the `theme()` function. 
This function is extremely powerful and initially it can be daunting to navigate the documentation. 

Fortunately there are many sources for help:

- you can search the help documentation from your console using `?theme`

- and you can check out the [ggplot2 website](https://ggplot2.tidyverse.org/reference/theme.html)

- if you have a specific issue, you can navigate to the website [StackOverflow](https://stackoverflow.com/) and search some key words. Say you are trying to change the font of the text in your plot, you can search "ggplot theme font family"

Let's take a look together:

type `?theme` on your console. In the help section of RStudio, the documentation of the function theme pops up.

Scroll down to the section  "Arguments": here you can find listed all the function arguments that you can use to customize your plot. As you can see a lot of the arguments are quite intuitive, making it easier to undertand what you may be looking for, e.g. `legend.position` and `axis.text.y` etc etc...

Once you find the arguments you need, remember to make your of the autocomplete function from RStudio

e.g. start typing in your R file `theme(legend)` and then hit the key TAB, you will see a number of options popping up. This greatly helps because you don't have to memorize all the arguments name.


**A USEFUL RULE OF THUMBS: **

- If you want to customize some *text elements* of your graph you will need to match the selected argument with `element_text()`

> example theme(axis.title = element_text(....))

- If you want to customize some *line elements* of your graph you will need to match the selected argument with `element_line()`

> example theme(axis.line = element_line(....))

- If you want to customize some elements of the plot or legend backround of your graph you will need to match the selected argument with `element_rect()`

> example theme(panel.background = element_rect(....))

- If you want to remove some elements from your graph you will need to match the selected argument with `element_blank()`

> example theme(panel.grid = element_blank(....))



Let's see an example of a graph- making workflow: 

1. Decide what to put in a plot and compose it 

1. Add some pre-defined theme using the function `theme_` 

1. Add titles and axis labels

1. Customize the plot appearance using the function `theme()`




```{r plot_themes}
basic_scatterplot <- NHANES %>% 
    ggplot(aes(x = Height, y = Weight, colour = Age)) +
    # use alpha (transparency) since there are so many dots
    geom_point(alpha = 0.5) +
    facet_grid(cols = vars(Gender)) +
    scale_color_viridis_c()

# Some pre-defined themes
basic_scatterplot + theme_bw()
basic_scatterplot + theme_minimal()
basic_scatterplot + theme_classic()
# Adding labels
basic_scatterplot +
    labs(title = "Height and weight by age and gender",
         x = "Height (cm)",
         y = "Weight (kg)")
# Edit specific plot items
basic_scatterplot +
    # theme is good at warning you if something isn't right
    # See ?theme for a full list of possible options
    theme(
        # background items must use element_rect
        # panel is the base/bottom layer that all other layers add to
        panel.background = element_rect(fill = "pink"),
        # strip is the section showing the facets
        strip.background = element_rect(fill = "red"),
        # line items must use element_line
        # axis is, well the axis
        axis.line = element_line(colour = "grey50", size = 0.5),
        # text items must use element_text
        # legend is the key when using fill, colour, size, etc
        legend.text = element_text(family = "sans", size = 20),
        #use element text to change the angle and position of axis           #labels
        #the function hjust controls the vertical alignment of labels
        axis.text.x = element_text(angle = 45, hjust = 1),
        # use element_blank to remove
        axis.ticks = element_blank()
    )
```



## Panels: join multiple figures together

Often we produce single figure that will then be assembled in one panel. 
This may happen due to journal requirements (e.g. limited number of figures 
allowed in a manuscript), or because different figures are conceptually linked 
and we want to show them close to one another.

The package `patchwork` offers some powerful and nifty functions to join figures in panels with a few lines of code.

First, if you have not done already make sure you have installed the 
package `patchwork`

```{r, eval = FALSE}
library(patchwork)
```


Let's say that we want to compare the look of the same plot "wearing" different 
themes. We will use the scatterplot we created earlier

```{r}
basic_scatterplot <- NHANES %>% 
    ggplot(aes(x = Height, y = Weight, colour = Age)) +
    # use alpha (transparency) since there are so many dots
    geom_point(alpha = 0.5) +
    facet_grid(cols = vars(Gender)) +
    scale_color_viridis_c()
```

First let's create the figure basic scatterplot adding a the `theme_minimal()` 
function and the title "Theme minimal" using the function `ggtitle()`
```{r}
basic_scatterplot_minimal <- basic_scatterplot +
  ggtitle("Theme minimal") +
  theme_minimal()
```

Same procedure but using the `theme_classic()` function and the title
"Theme classic"
```{r}
basic_scatterplot_classic <- basic_scatterplot +
  ggtitle("Theme classic") +
  theme_classic()
```

Now let's join these figures in a panel and compare them:

```{r}
basic_scatterplot_minimal + basic_scatterplot_classic 
```



We can also join more than two graphs in the same panel.
Let's create three simple graphs to use as an example
```{r}
plot1 <- ggplot(NHANES, aes(x = Height, y = BPSysAve, colour = Age)) +
    geom_point() +
    theme_minimal() +
    scale_colour_viridis_c()
   
plot2 <- ggplot(NHANES, aes(x = Height, y = Weight, colour = Age)) +
    geom_point() +
    theme_minimal() +
    scale_colour_viridis_c()


plot3 <- ggplot(NHANES, aes(x = Gender, fill = Education)) +
    geom_bar(position = position_dodge()) +
    facet_grid(cols = vars(Diabetes)) +
    scale_fill_brewer()
```


First, let's try join these plots together as we did before
```{r}
plot1 + plot2 + plot3
```

The result is difficult to read, because unless given further instructions `patchwork` tries and organizes plots in the same row.  Fortunately we can customize how graphs are dislayed:



```{r}
# | means: take plot1 and plot2 and put them in the same row
# / means: take plot 3 and put it into a separate row

(plot1 | plot2) / plot3
```

Better! Now plot3 is given more space in the bottom part of the panel, and everything is much easier to read.

However, the plot1 and plot2 share the same legend which now is repeated twice.

Using the function `plot_layout()` and its argument `guides` we can adjust that:
```{r}
((plot1 | plot2) / plot3) + 
  plot_layout(guides = 'collect')
```


Finally, we may want to add a title to our panel, and letters or numbers to the single figures in the panel, so we can refer to them separately when describing 
them in text

```{r}
((plot1 | plot2) / plot3) + 
  plot_layout(guides = 'collect') +
  plot_annotation(
    #adds a title to the panel
    title = "Colourful plots for my manuscript",
    #specifies what system to use to indicate plots in a panel
    tag_levels = "A"
  )
```


## Exercise: Joining figures in panels.

Choose figures from the past execises and join them in one panel with patchwork adding tags.

Example
```{r, eval= FALSE}
plot1 + plot2 + plot_annotation(____)
```

**Extra:**
Join three or more figures and experiment how to arrange them using the function `plot_layout(ncol)` , `plot_layout(ncol)` or the operators | and /

## Final exercise: Putting it all together

Time: Until end of session

<!-- In the `exercises-visualizing.R` script, add a new Section for this exercise. -->
Then using the NHANES dataset, try to replicate these figures. You very likely
won't be able to get exactly the same, but try to get close! :smile:

By the end of this you will try to get the data from `NHANES` to look like the
below figures. We don't expect you to get it exactly, but try to get close to
them. By (getting relatively close to) replicating these figures, you will have
achieved our learning expectations.

```{r final-exercise-plot1, warning=FALSE, message=FALSE, echo=FALSE}
NHANES %>% 
    filter(!is.na(Diabetes)) %>% 
    ggplot(aes(y = BMI, x = Poverty, colour = Diabetes)) +
    geom_point(alpha = 0.4) +
    geom_smooth(size = 2, method = "gam") +
    facet_grid(cols = vars(SurveyYr), rows = vars(Gender)) +
    theme_classic() +
    scale_color_viridis_d(end = 0.8) +
    theme(strip.background = element_blank(),
          panel.background = element_rect(fill = "grey95"),
          axis.line = element_line(colour = "grey80"))
```

```{r final-exercise-plot2, warning=FALSE, message=FALSE, echo=FALSE}
NHANES %>% 
    filter(!is.na(Diabetes), !is.na(Education)) %>% 
    ggplot(aes(x = Education, colour = Diabetes, y = TotChol)) +
    geom_boxplot(fill = "grey90", outlier.size = 0.5, size = 0.75) +
    theme_minimal() +
    scale_color_brewer(type = "qual") +
    labs(y = "Total Cholesterol") +
    facet_grid(cols = vars(Gender)) +
    coord_flip()
```

```{r final-exercise-plot1, echo=FALSE}
```

```{r final-exercise-plot2, echo=FALSE}
```
