---
title: "First steps in creating R packages"
author:
  - "Luke W. Johnston"
date: "2020-05-08"
categories:
  - Intermediate
  - R
  - Code-along
tags:
  - package development
  - R packages
  - usethis
  - devtools
  - workflow management
  - reproducible data analysis
---

> NOTE: This is still in development.

```{r setup, echo=FALSE}
source(here::here("R/utils-for-content.R"))
knitr::opts_chunk$set(eval = FALSE)
```

```{r, eval=FALSE}
install.packages(c("devtools", "usethis"))
```

### Objectives

1. To see that making packages is not as difficult as it seems.
1. To learn about and use the tools to simplify and automate package development.
1. To provide the resources to expand your knowledge on package development.

**At the end of this session you will be able:**

1. Setup up the necessary folders and files for making a package.
1. Fill out some package description information (in the `DESCRIPTION` and `README.md` files).
1. Include some functions in your package (in the `R/` folder).
1. Install your package.

## Resources for learning and help

[Here for resources](/more-info/#resources)

## List of functions and files covered

**Functions**:

- `usethis::create_package()`
- `usethis::use_r()`
- `usethis::use_readme_md()`
- `usethis::use_mit_license()`
- `devtools::load_all()` (Ctrl-Shift-L)
- `devtools::build()` / `devtools::install()` (Ctrl-Shift-B)
- `usethis::use_usethis()`

**Files**:

- `DESCRIPTION`
- `NAMESPACE`
- `R/`
- `README.md`
- `LICENSE`

## Creating and naming a package

First thing is first: An R package is simply a collection of R functions saved
on your computer that make it easy to access by using the `library()` command.
When you make an R package, it does *not* get uploaded online, it does *not* mean
other people can use it right away. At the beginning, the R package is only available
to you and to your computer. That's it. So why would you want to create a package?
Well, it can save time later with other data analyses since you can access your
functions across files by using `library(<yourpackagename>)`. A package bundles
related code together, simplifying your life! Alright, that's done with, next step!

Creating the initial setup of files and folders necessary for an R package is 
incredibly easy with the usethis package! But! Before we begin, it's a good idea
to think about what the name of the package should be. You can name it whatever
you want, as long as it is not "base", "stats", "graphics", "utils", "tools",
"methods", or any other default base R package. Generally, name it something that
is pretty unique or specific to you. For this session, let's call it by your
initials (or mine "lwj") and add "helpers"... so:

```{r, eval=FALSE}
usethis::create_package("lwj.helpers")
```

```{r, echo=FALSE}
pkg_name <- file.path(tempdir(), "lwj.helpers")
usethis::create_package(pkg_name)
```

A bunch of things are shown on your console. These are just some information
that usethis is telling you that it has done. Otherwise, you now have the beginning
of an R package! Pretty easy eh??

## RStudio project options

Before we get into the package details more, let's make some edits to the project
options. We'll change the "General" options to all "No" and check that the "Build
Tools" has correctly set the project as a package.

## Files and folders

There are many things that make an R package, well, an R package. One of them is
the folder and file structure. All R packages must have certain files and folders
in order to properly install. 

**Required**:

- `R/`: This folder contains the code that is available when the package is
installed. R files should only be in this folder, hence why it is called the
`R/` folder! Covered more below.
- `DESCRIPTION`: Contains all the necessary information about the package, such
as author, version number, name, brief description, and the package dependencies. 
This file is more of the "machine-readable" metadata of the package. Covered more
below.
- `NAMESPACE`: Tells R which functions are available to use from the package. 
At first, this file contains a command to list all functions that are in the `R/` 
folder. We won't cover this file too much during this session.

**Optional, but very helpful**:

- `.Rbuildignore`: This file let's R know what to ignore when you install your
package (for instance, the `.Rproj` file).
- The `.Rproj` file: Let's RStudio know that the directory is an [RStudio
Project](https://support.rstudio.com/hc/en-us/articles/200526207-Using-Projects)
and that it is a package. Adds some package development functionality to RStudio.

## Package metadata

As mentioned above, there are two main files for providing package "metadata".
Firstly, there is the `DESCRIPTION`, which is required for making a package.
There are several data-fields in the file:

```{r, echo=FALSE}
withr::with_dir(pkg_name, desc::desc())
```

The important data-fields are the `Title`, `Authors`, `Description`, and
`License` (get to this more later). Ok, let's open up the file and make the edits.

> **Exercise** (3-5 min): Open the `DESCRIPTION` file and change the title, author,
and description fields.

The next optional, but important file to have as "human-readable metadata" is the 
`README.md` file. Right now it doesn't exist. But we can create it using:

```{r, eval=FALSE}
usethis::use_readme_md()
```

```{r, echo=FALSE}
withr::with_dir(pkg_name, {
    capture.output(usethis::proj_set(pkg_name))
    usethis::use_readme_md()
})
```

The file looks like this:

```{r, comment="", echo=FALSE}
cat(readr::read_lines(file.path(pkg_name, "README.md")), sep = "\n")
# cat(file = file.path(pkg_name, "README.md"))
```

The `README.md` file is a [markdown](https://rmarkdown.rstudio.com/authoring_basics.html) 
file. We covered [markdown in a previous session](intro-rmarkdown.html).
Generally, this file is a brief documentation containing a high level summary of
what the package does, how to install it, and sometimes a simple example of how
to use it. It should be written and targetted to other (potential) users, or
yourself 9 months in the future (you'll thank yourself). Ok, this file right now 
has some things that need to be changed. Let's edit it.

> **Exercise** (5 min): Write a (fake) goal and description of the package in the
`README.md` file. Remove the text in the "Installation" section for now. Delete
the "Example" section.

Next we want to add the license. The license is really important, even if you 
never put your package online. The license gives information about how other people
can use and modify your code. We won't cover the legal details about licensing,
but for now, I would recommend using an [MIT License](https://en.wikipedia.org/wiki/MIT_License):

```{r, eval=FALSE}
# usethis::use_mit_license("<your name>")
usethis::use_mit_license("Luke Johnston")
```

```{r, echo=FALSE}
withr::with_dir(pkg_name, {
    capture.output(usethis::proj_set(pkg_name))
    usethis::use_mit_license("Luke Johnston")
})
```

The contents of the LICENSE file are:

```{r, comment="", echo=FALSE}
cat(readr::read_lines(file.path(pkg_name, "LICENSE")), sep = "\n")
```

## Adding R code

Now for the main reason you have created a package! To put your R code in it so
you can use the code easily in other scripts! To create an R script in the package,
use:

```{r, eval=FALSE}
# usethis::use_r("<filename>")
usethis::use_r("addition")
```

```{r, echo=FALSE}
# usethis::use_r("<filename>")
withr::with_dir(pkg_name, {
    capture.output(usethis::proj_set(pkg_name))
    usethis::use_r("addition")
})
```

An R script will open in RStudio. Now, let's add a function to it. We'll use a
very simple function in file:

```{r}
addition <- function(x, y) {
    x + y
}
```

Now save. To test that it works, now we run the next command:

```{r, eval=FALSE}
devtools::load_all()
```

Or type "Ctrl-Shift-L". In the console, test if the function works:

```{r, comment="#>"}
addition(2, 2)
```

It should work! We'll add some more functions and play around with the R scripts.

## Building and installing

Now, the final step! Let's build and install the package! When you feel your package
is ready, time to install it for you to use with `library()`! And you can install it
by using...:

```{r, eval=FALSE}
devtools::install()
```

Or by typing "Ctrl-Shift-B". It does some quick checks, and if all goes well, you
now have installed your package!

## Other useful functions

If you end up developing R packages often, I would recommend running this command:

```{r, eval=FALSE}
usethis::use_usethis()
```

This let's you add some code to automatically load when you open up R, to make it
easier to continue your package development!

<!--
> NOTE: Try to incorporate this content in or maybe make a new session out of it.
### Objectives

1. To learn how functions are structured and created.
1. To learn and apply the typical workflow for creating functions in packages
(and in general).
1. To properly document and name functions so you don't forget what they do! 

**At the end of this session you will be able:**

- Create a function with several arguments.
- Add and fill out roxygen documentation to the function.
    - Using the "Code"->"Insert Roxygen Skeleton" (`Ctrl-Shift-Alt-R`)
    - Then creating the documentation using `devtools::
- Cycle between testing the function out in the `@example` section and developing 
the function.
    - Using `devtools::load_all()` (`Ctrl-Shift-L`).
    - Run the code in the `@example` by typing `Ctrl-Enter`.
    - Repeat until the function does what you want!

**Ultimately**, I hope you will try to create a function from your own code by
the end of this session!

## Resources for learning and help

- Generally, the [R Packages](http://r-pkgs.had.co.nz/) book is great for
anything related to making packages.
- [`R/` chapter in R Packages](http://r-pkgs.had.co.nz/r.html) book (online and
free).
- [Documenting functions chapter](http://r-pkgs.had.co.nz/man.html) in R
Packages book.
- [Functions chapter in R for Data Science](https://r4ds.had.co.nz/functions.html)
book, also online and free.
- [Intro to functions](https://swcarpentry.github.io/r-novice-inflammation/02-func-R/)
from [Software Carpentry](https://software-carpentry.org/).
- [STAT545](http://stat545.com/block011_write-your-own-function-01.html)
on writing functions from University of British Columbia, Canada.
- Very advanced (for those interested) details about functions from [Advanced
R](http://adv-r.had.co.nz/Functions.html) book (again, online and free).

## Initial setup

First we need to create a package layout and create an R script. See the
[Package Creation](pkg-creation.html) session for more details on this.

1. Create a package using `usethis::create_package("<pkgname>")`
1. Add an R script using `usethis::use_r("<scriptname>.R")`.

## All actions in R are functions

The `+` is a function, `mean()` is a function, `[]` is a function... everything 
that does something is called a function in R. So to add 1 with 1:

```{r}
1 + 1
```

... is a function that takes 1 and adds 1 to it. It is actually a short form for:

```{r}
`+`(1, 1)
```

When creating a function, there is always the basic structure of:

1. Name of the function (e.g. `mean`).
2. The function call using `function()` being assigned `<-` to the name. This
tells R that the name is a function object... that it does some action.
3. The arguments within the function call, `function(arg1, arg2, arg3)`. These
are the options given to the function (e.g. `sum(arg1, arg2)`).
4. The body of the function, that takes the arguments, if any, does some action,
and finishes by outputting some result or object at the end (using `return()`).

There is no minimum or maximum number of arguments you can provide for a function.
E.g. you can have zero arguments or you can have 100s. To keep things sane, try
to keep the number of arguments low, like not more than 4 or 5.

So, the structure is:

```{r, eval=FALSE}
name <- function(arg1, arg2) {
    # body of function
    return(output)
}
```

... and an example:

```{r}
add_nums <- function(num1, num2) {
    added <- num1 + num2
    return(added)
}
```

You can use the new function by running the above code and writing out your new 
function, with arguments to give it.

```{r}
add_nums(1, 2)
```

The function name is fairly good... `add_nums` can be read as "add numbers". But
we need to also add some formal documentation to the function. Using the "Insert
Roxygen Skeleton" in the "Code" menu list (or by typing `Ctrl-Shift-Alt-R`) you 
can add template documentation right above the function. It looks like:

```{r}
#' Title
#'
#' @param num1 
#' @param num2 
#'
#' @return
#' @export
#'
#' @examples
add_nums <- function(num1, num2) {
    added <- num1 + num2
    return(added)
}
```

In the `Title` area, this is where you type out a brief sentence or several words 
that describe the function. Creating a new paragraph below this line allows you 
to add a more detailed description. The other items are:

- `@param num` lines are to describe what each argument is for.
- `@return` describes what output the function provides. Is it a data.frame? A plot? 
What else does the output give?
- `@export` this is to tell R that this function should be accessible to the user
of your package. Keep it in for now.
- `@examples` lines below this are used to show examples of how to use the function.
This is also the area where you write and test that the function does what you want.
You can run code here as if it was regular R code and not a commented out code by
using `Ctrl-Enter`.

```{r}
#' Add two numbers together.
#'
#' This is just an example function to show how to create one.
#'
#' @param num1 A number here.
#' @param num2 A number here.
#'
#' @return Returns the sum of the two numbers.
#' @export
#'
#' @examples
#'
#' add_nums(1, 1)
add_nums <- function(num1, num2) {
    added <- num1 + num2
    return(added)
}
```

Now, when we run `devtools::document()` (or `Ctrl-Shift-D`), a file will be added
to the `man/` folder. Now, when you type out `?add_nums` in the console, the help
documentation will pop up on the "Help" tab.

Ok, let's get to something a bit more interesting. A common thing that people do
(at least I do) is to create a similar plot on different variables and datasets.
So this is a great example of using a function to simplify your code. We can also
cover... package/function dependencies! Since we will use ggplot2 to make the plot,
we need some way to tell R that the functions come from ggplot2... It's **very bad
practice** to use `library()` in your function and in your package. A function from
usethis comes to the rescue! Use `usethis::use_package("ggplot2")` in the
console. Some text will appear saying that "ggplot2 has been added to DESCRIPTION" 
and that "please refer to functions using ggplot2::fun". So let's do that!

```{r}
scatter_plot <- function(data, xvar, yvar) {
    graph <- ggplot2::ggplot(data = data, ggplot2::aes(x = xvar, y = yvar)) +
        ggplot2::geom_point()
    return(graph)
}
```

Ok, we have the base for making a scatter plot. But! There are few things to talk
about here first. First, ggplot2 will be confused by the `x = xvar` since it will
think[^nse] you are asking for the `xvar` column in the dataset. So, we need to change
`aes` to `aes_string` to force ggplot2 to read `xvar` as a character string that
is the name of the column you want to plot. Next, it is useful in many cases to 
put a dot before your function arguments to differentiate your function arguments
from other R objects. Let's also add the Roxygen documentation. So:

[^nse]: The reason it won't work is because ggplot2 is using what is called
non-standard evaluation (NSE; check out
[here](https://cran.r-project.org/web/packages/lazyeval/vignettes/lazyeval.html)
or [here](https://dplyr.tidyverse.org/articles/programming.html) for more
indepth look at what non-standard evaluation is). Because ggplot2 uses NSE, you
will have to do things slightly differently, hence using `aes_string`.

```{r}
#' Create a scatter plot of two variables.
#'
#' @param .data The dataset.
#' @param .xvar The x-axis variable.
#' @param .yvar The y-axis variable.
#'
#' @return A scatter plot.
#' @export
#'
#' @examples
#'
#' scatter_plot(swiss, "Education", "Agriculture")
scatter_plot <- function(.data, .xvar, .yvar) {
    graph <- ggplot2::ggplot(data = .data, ggplot2::aes_string(x = .xvar, y = .yvar)) +
        ggplot2::geom_point()
    return(graph)
}
```

```{r, echo=FALSE}
scatter_plot(swiss, "Education", "Agriculture")
```

Now we do `devtools::load_all()` (`Ctrl-Shift-L`) and run the code in the
example (`Ctrl-Enter`). A quick note: Running code in the `@examples` section *only*
works if you are in an R package project (the RStudio `.Rproj` file). If you are
not in an R package, you can instead include the code to run the function below
your new function, like so:

```{r, eval=FALSE}
#' Create a scatter plot of two variables.
#'
#' @param .data The dataset.
#' @param .xvar The x-axis variable.
#' @param .yvar The y-axis variable.
#'
#' @return A scatter plot.
#' @export
#'
#' @examples
#'
scatter_plot <- function(.data, .xvar, .yvar) {
    graph <- ggplot2::ggplot(data = .data, ggplot2::aes_string(x = .xvar, y = .yvar)) +
        ggplot2::geom_point()
    return(graph)
}
scatter_plot(swiss, "Education", "Agriculture")
```

Now, if we want to add some theme items, all graphs created from this function 
will get the new theme and appearance!

```{r}
#' Create a scatter plot of two variables.
#'
#' @param .data The dataset.
#' @param .xvar The x-axis variable.
#' @param .yvar The y-axis variable.
#'
#' @return A scatter plot.
#' @export
#'
#' @examples
#'
#' scatter_plot(swiss, "Education", "Agriculture")
scatter_plot <- function(.data, .xvar, .yvar) {
    graph <- ggplot2::ggplot(data = .data, ggplot2::aes_string(x = .xvar, y = .yvar)) +
        ggplot2::geom_point() +
        ggplot2::theme_minimal()
    return(graph)
}
```

```{r, echo=FALSE}
scatter_plot(swiss, "Education", "Agriculture")
```

If you want to make sure that who ever uses your function will not use a wrong 
argument, you can use "defensive programming" via the `stopifnot()` function.
This forces the code to only work if `xvar` and `yvar` are character (e.g. `"this"`)
argument.

```{r}
#' Create a scatter plot of two variables.
#'
#' @param .data The dataset.
#' @param .xvar The x-axis variable.
#' @param .yvar The y-axis variable.
#'
#' @return A scatter plot.
#' @export
#'
#' @examples
#'
#' scatter_plot(swiss, "Education", "Agriculture")
scatter_plot <- function(.data, .xvar, .yvar) {
    stopifnot(is.character(.xvar), is.character(.yvar))
    graph <- ggplot2::ggplot(data = .data, ggplot2::aes_string(x = .xvar, y = .yvar)) +
        ggplot2::geom_point() +
        ggplot2::theme_minimal()
    return(graph)
}
```

## Exercise: Make your own function!

Using this workflow, try to create, document, and test your own function! If you
have some code that you already use repeatedly by copy and pasting, try to convert
that code into a function. If you don't have your own code, try converting these
pieces of code into their own function:

```{r, eval=FALSE}
# Code chunk 1:
one_species <- iris[iris['Species'] == "setosa", ]
sepal_one_species <- one_species[c("Sepal.Length", "Sepal.Width")]
means_sepal_one_species <- sapply(sepal_one_species, mean)
means_sepal_one_species
```

## Other notes

There are a few other things to consider. In R there are different "methods" of
functions. This is way above what is necessary for this session, but if you are
curious [this website](http://adv-r.had.co.nz/S3.html) has a great explanation
of the different methods (e.g. S3 methods). Be warned, the website is fairly
advanced!

You can always look at the contents of all functions in R. So an example of an
S3 function:

```{r}
# Generic S3
print
# Printing for data.frames
print.data.frame
```
-->
