---
title: "Data visualization with ggplot2, part 2"
author:
  - "Luke W. Johnston"
  - "Sara Tomiolo"
date: "2020-01-31"
description: >
  Dig deeper into making the most of making data visualizations with ggplot2.
  We'll show you how to plot multiple panels within a single plot, how to append
  multiple plots together, and how to make your plot more visually appealing.
categories:
  - Intermediate-Beginner
  - R
  - Code-along
tags:
  - ggplot2
  - data visualization
  - colour palettes
  - patchwork
---

```{r setup, echo=FALSE}
source(here::here("R/utils-for-content.R"))
```

# Session details

## Preparation

Please install these packages:

```{r pkg-install, eval=FALSE}
install.packages(ggplot2)
install.packages(patchwork)
install.packages(viridis)
install.packages(NHANES)
```

## Objectives

By the end of this session you will know how to:

1. Apply facetting functions to include multiple "panels" on a single plot.
1. Consider the challenges with using colours and to make use of appropriate colour palettes.
1. Modify the theme settings to make visually appealling plots.
1. Discover and use new theme settings to modify nearly all aspects of a plot.

**Resources** for continuing learning and help are available on
the [More Info](/more-info/#resources) page.

# Getting more out of ggplot2

## First steps: Project setup and exploring the data

Like most teaching sessions, we'll start with creating an R project. 
R projects are incredibly useful tools and we strongly recommend you use them.
Create a new project, name it "more-ggplot2". 
Once the R project has been opened, create two new files and save them as 
`code-along.R` and `exercises.R`.

Next, let's get familiar with the data we'll be using for this session, 
which is the `NHANES` dataset. 
Open up the `code-along.R` file and we'll add these to the file:

```{r load-packages-see-data}
# Import the packages we'll use
library(NHANES)
library(ggplot2)
library(patchwork)
library(viridis)

# Content
NHANES
```

## Basic structure of using ggplot2

As a review, we'll cover the basic structure of ggplot2.
ggplot2 uses the "Grammar of Graphics" (gg). This is a powerful approach to
creating plots as it provides a consistent way of telling ggplot2 what to
do. There are at least three aspects to using ggplot2 that relate to the grammar:

- Aesthetics, `aes()`: How data should be mapped to the plot. Includes what to
put on x axis, on the y axis, colours, size, etc.
- Geometries, `geom_`: The visual representation of the data, as a layer. This
tells ggplot2 how to show the aesthetics. Includes points, lines, boxes, etc.
- Scales, `scale_`: Allows more direct control of how data values are visually presented.
For instance, `scale_x_continuous()` allows setting visual properties of
continuous values on the x-axis.
- Themes, `theme_` or `theme()`: How the plot should look like. Using this
function we can customize several components of a graph, such as axis, lines,
grids, plot background, position, size and font of the axis text and labels
- Title and labels can be added using the functions `ggtitle()` and `labs()`
respectively

The aesthetics and the geometries are fundamental to create a graph, whereas
scales, themes, and labels are features that can be stacked in successive layers
to improve the looks and the clarity of a graph.

**Tip:** To maximise the power of ggplot2, make heavy use of autocompletion. You
can do this by typing, for instance, `geom_` and then hitting the TAB key to see
a list of all the geoms. Or after typing `theme()`, hit TAB to see all the
options inside theme.


### Using colours to convey information

You may have used colours before when making a plot.
Adding colours isn't just for visual appeal. 
It can also convey more information about your data and results.
Colours can be used to represent discrete groups like male or female
or they can be used to represent a continuity of values, like higher or lower 
(in e.g. elevation as in maps).
Let's try it out:

```{r plot-setup}
# Two continuous and one discrete
# Scatter plot
two_continuous <- ggplot(NHANES, aes(x = Height, y = Weight, colour = Gender)) +
  geom_point()
two_continuous

# One continuous and two discrete
two_discrete <- ggplot(NHANES, aes(x = Gender, y = BMI, colour = Diabetes)) +
  geom_boxplot()
two_discrete
```

However, something to consider when using colours is that not everyone sees how you see.
Colour blindness is common in the population, 
and red-green colour blindness in particular affects 8% of men and 0.5% of women. 
Making your graph more accessible to people affected by colour blindness 
will also usually improve the interpretability of your graphs for all people. 
For more detail on how colours look to those with colour-blindness, 
see this [documentation](https://cran.r-project.org/web/packages/viridis/vignettes/intro-to-viridis.html) 
from the [viridis] package. 
The viridis colour scheme and R package was specifically designed to represent 
data well to all colour visions. 
There is also a really good, 
informative [talk on YouTube](https://www.youtube.com/watch?v=xAoljeRJ3lU) on this topic.

[viridis]: https://cran.r-project.org/packages=viridis

When using colours, take time to think about what you are trying to convey in
your figure and how your choice of colours will be interpreted. You can use
built-in colour schemes, or set your own. Let's stick to using builtin ones.
There are two, the viridis and the [ColorBrewer] colour schemes. Both are well
designed and are colour-blind friendly.

[ColorBrewer]: http://colorbrewer2.org/

Let's try out viridis using the previous plots we made:

```{r using-viridis-discrete}
# Two continuous and one discrete
# Default colour scheme
two_continuous +
  scale_color_viridis_d()

# Magma colour scheme
two_continuous +
  scale_color_viridis_d(option = "magma")

# One continuous and two discrete
# Plasma colour scheme
two_discrete +
  scale_color_viridis_d(option = "plasma")

two_discrete +
  scale_color_viridis_d()
```

Viridis is a colour scheme, with a range of colours.
So you can also use the colour scheme with continuous variables.
But you'll need to use the `_c()` form of viridis: `scale_color_viridis_c()`.

```{r using-viridis-continuous}
three_continuous <- ggplot(NHANES, aes(x = Height, y = Weight, colour = Age)) +
  geom_point()

# Viridis colour scheme
three_continuous +
  scale_color_viridis_c()

# Inferno colour scheme
three_continuous +
  scale_color_viridis_c(option = "inferno")

# Magma colour scheme
three_continuous +
  scale_color_viridis_c(option = "magma")

# Can change ending or beginning colour
# Range goes from 0 for begin to 1 for end
three_continuous +
  scale_color_viridis_c(end = 0.8)

three_continuous +
  scale_color_viridis_c(begin = 0.2)
```

The ColorBrewer also has some good colour blind-friendly palettes.
See all the colour palettes for the Brewer palette using
`RColorBrewer::display.brewer.all()`.
Check out `scale_color_brewer()` for discrete data and
`scale_color_distiller()` for continuous data.
We won't go over this, but here's some code to get started:

```{r using-brewer, eval=FALSE}
two_discrete +
  scale_color_brewer(palette = "Set1")

two_discrete +
  scale_color_brewer(palette = "Dark2")

# Use qual for qualitative data
two_discrete +
  scale_color_brewer(type = "qual", palette = "Dark2")

three_continuous +
  scale_color_distiller()

three_continuous +
  scale_color_distiller(palette = "Spectral")
```

## Facetting: adding even more variables to a plot

Facetting in ggplot2 is an incredibly powerful way to quickly visualize multiple
perspectives of a dataset.
Facets are groupings of plots by a discrete variable.
There are two functions in ggplot2 for facetting: 

- `facet_grid()`: Have on variable split by column (horizontally) and another one
split by row (vertically).
- `facet_wrap()` takes a variable and splits it up and arranges it for "n" number 
of columns, after which it will wrap to the next row.

Both facets must be given the `vars()` function when the variable is provided.
Let's try it out!

```{r plot-facet}
two_discrete +
    facet_grid(cols = vars(SurveyYr))

two_discrete +
    # Rows means to have them stacked vertically
    facet_grid(cols = vars(SurveyYr), rows = vars(HomeOwn))

two_continuous +
    # Rows means to have them stacked vertically
    facet_grid(cols = vars(SurveyYr), rows = vars(HomeOwn))
```

If you know how to convert the data to long format using `pivot_longer()`,
you can really take advantage of facets. 
However, we won't be covering that in this session.

## Exercise: Try out facets with the colour palettes

Time: 10-15 min

In the `exercises.R` script,
copy the code below into the script. 
Then do the following:

1. Look in the NHANES dataset and choose two continuous and two discrete variables.
1. Put the two continuous variables as the `x` and `y` in `aes()`.
1. Put one of the discrete variables as the `colour`.
1. Choose either `scale_color_viridis_d()` or `scale_color_brewer()` for the palette.
1. For `facet_grid()` choose to use either `rows = ` or `cols =` for the second 
discrete variable. Don't forget to use `vars()` when giving the variable name.

Do the same thing for the other plots, 
except choose three continuous variables for 2. and
three discrete variables for 3. 
For 3., use `facet_wrap()` instead to try it out.

```{r exercise-facet, eval=FALSE}
# 1. Two continuous, discrete, and one facet.
ggplot(NHANES, aes(x = ___, y = ___, colour = ___)) +
    geom_point() +
    scale____ +
    facet_grid(___)

# 2. Three continuous
ggplot(NHANES, aes(x = ___, y = ___, colour = ___)) +
    geom_point() +
    scale____ +
    facet_grid(___)

# 3. Three discrete
ggplot(NHANES, aes(x = ___, fill = ___)) +
    geom_bar(position = position_dodge()) +
    scale____ +
    facet_wrap(vars(___), ncol = ___)
```

## Adding together multiple plots

Often we produce multiple plots that will then be assembled in one figure. 
This may happen due to journal requirements (e.g. limited number of figures 
allowed in a manuscript), or because different figures are conceptually linked 
and we want to show them close to one another.

The package patchwork offers some powerful and 
nifty functions to join figures in panels with a few lines of code.

First, open up the `code-along.R` script again.
Let's say that we want to compare the look of the same plot "wearing" different themes. 
We'll create a scatterplot:

```{r}
basic_scatterplot <- ggplot(NHANES, aes(x = Height, y = Weight, colour = Age)) +
    # Use alpha (transparency) since there are so many dots
    geom_point(alpha = 0.5) +
    facet_grid(cols = vars(Gender)) +
    scale_color_viridis_c()
```

Now, we'll add use the `theme_minimal()` and add an appropriate title with `ggtitle()`.

```{r}
basic_scatterplot_minimal <- basic_scatterplot +
  ggtitle("Theme minimal") +
  theme_minimal()
```

We'll do the same thing again but with `theme_classic()`.

```{r}
basic_scatterplot_classic <- basic_scatterplot +
  ggtitle("Theme classic") +
  theme_classic()
```

Now let's join these figures in a panel and compare them:

```{r}
basic_scatterplot_minimal + basic_scatterplot_classic 
```

Cool eh?

But this is too simple an example. 
Let's do something a bit more realistic and this time add together three plots.
Let's create three simple graphs that each have different underlying `aes()` and `geom_` layer.

```{r setup-plots-patchwork}
plot1 <- ggplot(NHANES, aes(x = BMI, y = BPSysAve, colour = Age)) +
    geom_point() +
    theme_minimal() +
    scale_colour_viridis_c()
   
plot2 <- ggplot(NHANES, aes(x = BMI, y = TotChol, colour = Age)) +
    geom_point() +
    theme_minimal() +
    scale_colour_viridis_c()

plot3 <- ggplot(NHANES, aes(x = Gender, fill = Education)) +
    geom_bar(position = position_dodge()) +
    facet_grid(cols = vars(Diabetes)) +
    theme_minimal() +
    scale_fill_brewer()
```

First, let's try join these plots together as we did before.

```{r arrange-plots-1}
plot1 + plot2 + plot3
```

Hmm... The result is a bit difficult to read.
patchwork by default adds plots horizontal on one row.
Fortunately we can customize how graphs are put together:

```{r arrange-plots-2}
# | means: take plot1 and plot2 and put them in the same row
# / means: take plot 3 and put it into a separate row
(plot1 | plot2) / plot3
```

Better! Now `plot3` is given more space in the bottom part of the figure, and
everything is much easier to read.
However, `plot1` and `plot2` share the same legend which now is repeated twice.
We can use the function `plot_layout()` and its argument `guides` to adjust that:

```{r guide-collect}
((plot1 | plot2) / plot3) + 
  plot_layout(guides = 'collect')
```

Finally, we may want to add a title to our figure, 
and letters or numbers to the single plots in the figure, so we can refer to them separately when describing 
them in text

```{r}
((plot1 | plot2) / plot3) + 
  plot_layout(guides = 'collect') +
  plot_annotation(
    # Adds a title to the figure
    title = "Colourful plots for my manuscript",
    # Specifies what system to use to indicate plots in a figure
    tag_levels = "A"
  )
```

## Exercise: Joining together plots in a figure

Time: 10 min

Taking the plots from the previous execise,
join them together into one figure with patchwork.
Open the `exercise.R` file and do the following:

1. Put the previous plots into objects called `plot1`, `plot2`, etc.
1. Add plot tags as letters. See `?plot_annotation` and 
the documentation on `tag_levels` to see other types of tags. 
Use this code as a template.

```{r exercise-patchwork, eval= FALSE}
plot1 + plot2 + plot_annotation(____)
```

1. Arrange the plots vertically instead using `/`. 
You can also use the `plot_layout(ncol = 1)` to only have one column of plots.

## Themes: customize text, line, background and legends

There are so so many options to modify the figure, 
and they are all basically contained in the `theme()` function. 
This function is extremely powerful and 
initially it can be daunting to use and to navigate the documentation. 

Fortunately there are many sources for help:

- You can search the help documentation from your console using `?theme`
- You can check out the [ggplot2 website](https://ggplot2.tidyverse.org/reference/theme.html)
- If you have a specific issue, 
you can navigate to [StackOverflow](https://stackoverflow.com/) and search some key words. 
Say you are trying to change the font of the text in your plot, 
try searching for "ggplot theme font family".

Let's take a look together:

Type `?theme` in your console. 
In the help section of RStudio, 
the documentation of the function theme pops up.

Scroll down to the section "Arguments": 
here is listed all the function arguments that you can use to customize your plot. 
As you can see, a lot of the arguments are quite intuitive, 
making it relatively easy to learn and undertand what you may be looking for.
For instance, if you want to change the legend position,
the argument is `legend.position`. 
If you want to change the y axis text, the argument is `axis.text.y`.
And so on.

Once you find the arguments you need, make use of RStudio's auto-complete.
Let's try it with the legend position. 
Start typing out `theme(legend` and then hit TAB to see a list of possible arguments to use.
This helps out a lot because you don't have to memorize all the arguments name.

Each argument in `theme()` usually needs to take an `element_` type function.
The type of `element_` to use depends on the item you are wanting to modify.
So a rule of thumb is that the element name is linked to the name of what you want to change:

- To customize *text* elements of your graph,
you'll need to match the selected argument with `element_text()`:

  ```r
  theme(axis.title = element_text(....))
  ```

- To customize *line* elements of your graph,
you'll need to match the selected argument with `element_line()`:

  ```r
  theme(axis.line = element_line(....))
  ```

- To customize elements of the plot or legend background,
you'll need to match the selected argument with `element_rect()`:

  ```r
  theme(panel.background = element_rect(....))
  ```

- To remove some elements from your graph,
you'll need to match the selected argument with `element_blank()`:

  ```r
  theme(panel.grid = element_blank(....))
  ```
  
Let's try adding some theme items to our `basic_scatterplot`. 
Open up the `coe-along.R` file.

```{r plot-themes}
# Edit specific plot items
basic_scatterplot +
    # Theme is good at warning you if something isn't right
    # See ?theme for a full list of possible options
    theme(
        # Background items must use element_rect
        # Panel is the base/bottom layer that all other layers add to
        panel.background = element_rect(fill = "pink"),
        # Strip is the section showing the facets
        strip.background = element_rect(fill = "red"),
        # Line items must use element_line
        # Axis is, well the axis
        axis.line = element_line(colour = "grey50", size = 0.5),
        # Text items must use element_text
        # Legend is the key when using fill, colour, size, etc
        legend.text = element_text(family = "sans", size = 20),
        # use element_text to change the angle and position of axis labels
        # the function hjust controls the vertical alignment of labels
        axis.text.x = element_text(angle = 45, hjust = 1),
        # use element_blank to remove
        axis.ticks = element_blank()
    )
```

## Exercise: Change theme items of the plot

Time: 10 min

Open up the `exercises.R`. 
Using the approach to finding theme items in the help documentation,
make changes to at least three theme items.
Choose at least one `axis` item, one `panel` item, and one `legend` item.
Use the below code as a template to start:

```{r exercise-theme, eval=FALSE}
ggplot(NHANES, aes(x = Age, y = BMI)) +
    geom_point() +
    theme(
      axis.___ = element_
      panel.___ = element_
      legend.___ = element_
    )
```
