---
title: "Efficient coding when iterating"
author:
  - "Florian Privé"
  - "Luke W. Johnston"
date: "2020-01-31"
description: >
  Do you repeat yourself often when writing code? Do you make use of loops? Come
  to this session to learn how to write R code more efficiently and for some best
  practices to follow.
categories:
  - Beginner-Intermediate
  - R
  - Code-along
tags:
  - purrr
  - functional programming
  - iterations
---

```{r setup, echo=FALSE}
source(here::here("R/utils-for-content.R"))
```

# Session details

## Preparation

Please install these packages:

```{r pkg-install, eval=FALSE}
install.packages(tidyverse)
```

## Objectives

By the end of this session you will know how to:


**Resources** for continuing learning and help are available on
the [More Info](/more-info/#resources) page.

# Writing R code that iterates more efficiently

## First steps: Project setup and exploring the data

Like most teaching sessions, we'll start with creating an R project. 
R projects are incredibly useful tools and we strongly recommend you use them.
Create a new project, name it "iterating-with-purrr". 
Once the R project has been opened, create two new files and save them as 
`code-along.R` and `exercises.R`.

Next, let's get familiar with the data we'll be using for this session, 
which is the {{DATA}} dataset. 
Open up the `code-along.R` file and we'll add these to the file:

```{r load-packages-see-data}
# Import the packages we'll use
library(tidyverse)

# Content
# {{DATA}}
```

## SECTION

## Exercise: Try out ...

Time: 10-15 min

In the `exercises.R` script,
copy the code below into the script. 
Then do the following:

1. 
1. 

```{r exercise-1, eval=FALSE}
# ...
```

## NEXT SECTION

## Exercise: ...

Time: 10 min

Open the `exercise.R` file and do the following:

## NEXT SECTION

## Exercise: ...

Time: 10 min

Open up the `exercises.R`. 

```{r exercise-3, eval=FALSE}
# ...
```
