---
title: "2020-01-19-intermediate-dplyr"
author: 
- "Sara Tomiolo"
- "Luke Johnston"

date: "1/17/2020"
description: >
  Learn to reshape data in long or wide form, manage factors and dates using the packages dplyr, tidyr, lubridate, forcats and stringr.
categories:
  - Intermediate
  - R
  - Code-along
tags:
  - tidyr
  - dplyr
  - lubridate
  - data manipulation
  - dates
---


```{r setup, echo=FALSE}
source(here::here("R/utils-for-content.R"))
```

## Preparation

Please install the tidyverse package:

```{r, eval=FALSE}
install.packages("tidyverse")
library(tidyverse)
library(lubridate)
```

### Prerequisites

In this lesson, we assume a bit of familiarity with the basic functions of dplyr (especially `group_by()` and `summarize()`), which we covered in the [intro_dplyr lesson](https://au-cru.github.io/site/material/2019-11-15-intro-dplyr/) 


### Objectives

1. To learn to restructure your data between long and wide format with 
`pivot_longer()` and `pivot_wider()` functions from tidyr package.
2. Learn to combine `group_by()` and `summarize()` function with `pivot_longer()` and `pivot_wider()` functions.
3. Manage dates with lubridate functions.

    
**At the end of this session you will be able:**

1. To re-structure your data in long or wide format.
2. To reshape data by using grouping, summarizing and reshaping dataframes.
3. To reformat dates with lubridate

### Why learning to reformat and reshape data?

Often we receive databases that have been created by others, and in which information is organized in a "messy" way. Some other times, we simply need to reorder the structure of the data to be able to carry out certain data analysis or data visualization that require a specific data structure. 

The tidyverse, provides some very powerful packages (e.g. dplyr, tidyr) that can allow for complete reorganization of our data in a few lines of codes. Using these packages and their functions have several advantages:

- It makes our lives easier and our work faster because we can jump between different 
formats of our data in an agile and easy-to-track way;

- It makes our work reproducible.

### Wide and long formats

One of the most common operations that we may need to do is to change the way our data is organized from a long (few columns, many rows) to a wide (few rows, many columns) format.

The functions we want to use for these operations are `pivot_longer()` and `pivot_wider()` from the tidyr package (Those of you already familiar with this package may know these functions as `spread()` and `gather()`. Although these two functions still work, they are being slowly replaced by the `pivot_**` functions).

To show how these functions work, we are going to use the built-in dataframe `population`.

#### Familiarize with the dataframe

Using the function `View()` inspect the dataframe population, which shows the population census in different countries across multiple years.

We are going to load tidyverse

```{r}
library(tidyverse)
```


```{r}
View(population)
```


### `Pivot_wider()`
Say that we want to explore how the population varies across years in each country. 
To do this we may want to show each single country as a separate variable (i.e. column). 

This requires reshaping the data so that each single country now listed under the variable `country` 
becomes an individual variable (i.e. column), under which the population numbers, 
now listed under the variable `population`, are listed.

To do so, we will use the function `pivot_wider()` because we want to create 
a wider dataframe with more columns (the single countries), and fewer rows.

We will call this new dataframe `population_wide`.
In the function `pivot_wider()` we want to specify three arguments: 

- the dataframe to work on,
- the argument `names_from`: where we indicate the name of the "long" variable that we will use to create the new "wide" variables 
- the argument `values_from`: the values that we will use to fill the rows of the new variable we created

```{r}
population_wide <- 
  pivot_wider(population, names_from = country, values_from = population) 
```

We can run the exact same operation taking advantage of the pipe operator `%>%`
that we showed in the intro session. If the memory is foggy, imagine this operator as an
arrow that connects consecutive functions in a smooth way. 

When using the pipe operator we are indicating that the functions will be applied to the 
dataframe `population`, so we do not need to indicate that into the `pivot_wider()` function arguments.

```{r}
population_wide <- population %>% 
  pivot_wider(names_from = country, values_from = population) 
```


#### Exercise 1
Use the dataframe `us_rent_income` for this exercise. This dataframe shows the estimated value and the 90%
margin of error for yearly income and monthly rent of each state in the US.

In this case, you want to display the data in wide form to see how estimated income varies across states.
To do so, each state will become a variable (i.e. column) to which the estimated income will be associated

```{r, eval FALSE}
us_income_wide <- us_rent_income %>% 
  #because we are only interested in income and not in rent, we will extract the variable of interest from our database
  filter(variable == "income") %>% 
  #we select only the variables of interest to remain in our dataframe
  select(NAME, variable, estimate) %>% 
  pivot_wider(names_from = ______, values_from = _____)
```

### `Pivot_longer()`

We can also do the opposite: a shift from a dataframe in wide format to a dataframe in long format. 
Let's consider the `population_wide` dataframe that we just created as the starting point. 
Let's say that we want to reshape this data to a format in which all countries are listed under a single variable, e.g. `country_name`, and the population data are listed under a variable called `population_data`. For this operation, we need to use the function `pivot_longer()`

In this case, we want to specify:
- the argument `names_to`: the name of the new variable under which we want to group our columns, 
- the argument `values_to`: the name of the new variable under which we want to group the values of our variables.

We run this function...

```{r}
population_wide %>%
  pivot_longer(names_to = "Country", values_to = "Population")
```

... but we get an error message!

This is because we did not specify which columns to shift into a long format.

  > In `pivot_wider()`, if no argument is given to `id_cols` it defaults to select all columns. However, `pivot_longer` has no default for `cols`, and will throw an error if no argument is given.

We could select every column, with the tidyselect function `everything()`. This is not what want (..., but you should try it to see why!). 

We want to keep the row-wise relation between year and population, So we specify that the variable `year` is not to be involved in the reshaping.

```{r}
population_wide %>%
  pivot_longer(-year, names_to = "country_name", values_to = "population_data")
```


**Notice that** in the function `pivot_longer()` the arguments `names_to` and `values_to` require "" because we are providing names of variables that need to be created. 
Instead, in the function `pivot_wider()` we are giving the arguments `names_from` and `values_from` variables that already exist, hence  the "" are not needed.

#### Exercise

We will use again the `us_rent_income` dataframe for this exercise.
In this case we want the values of `estimate` and `moe` to be stacked in one single column called `value`, 
while the respective categories of `estimate` and `moe` should be described in a variable called `calculation`

```{r, eval = FALSE}
us_rent_income_long <- us_rent_income %>% 
  #this variable is a nuisance and is not giving information useful to us, so we remove it 
  select(-GEOID) %>% 
  #remember that the variable that will not be reshaped need to be omitted using the operator -
  #when there is more than one variable to omit, we combined the operators - with c()
  pivot_longer(-c(______, ______), names_to = ________, values_to = _________)
```



### Use `pivot_**` functions in combination with `group_by()` and `summarize()` functions

For this example we are going to use the built-in dataframe `storms`

Once again, explore the data using the function `View()`

```{r}
View(storms)
```

In this dataframe, information about storms are classified  in long format. 
Storms are categorized across years and information on wind, latitude, longitude and pressure 
are presented as columns.

Let's say that we want to see how mean wind speed vary across 
different storms (name and status) and across years. 

This means that we want to apply the function `pivot_longer()` to visualize each storm name as a separate variable.

However, before reshaping the dataframe in long form, we need to do some extra work. 
Take another look at the `storms` dataframe. Variables are measured multiple times within each year. 
So if we want to have a mean value of wind speed for each year, we need to use the functions `group_by()` and `summarize()` that we showed in the intro session. 

We are going to call this new dataframe `storm_wide`

First we will calculate the mean value of wind speed for each storm, across years 
and discriminating across storm status

```{r}
storms_wide <- storms %>% 
  #first we select the variable of interest to avoid an overcrowded dataframe
  select(name, year, wind, status) %>% 
  #second we tell R according to which criteria we want to summarize our data, i.e. we want to know the mean value of wind speed for each storm according to storm status and year
  group_by(name, year, status) %>% 
  summarize(mean_wind = mean(wind)) 
```

Now we can reshape the data into a wide format 
```{r}
storms_wide <- storms_wide %>% 
   pivot_wider(names_from = name, values_from = mean_wind)  
```

#### Exercise
Using the dataframe `us_rent_income` use the functions `group_by()` and `summarize()` in combination with
`pivot_longer()`, `pivot_wider()` to reshape a database in which different values of cut correspond to mean values of x, y, z (tip: you need to use both pivot_longer and pivot_wider!)

The reshaped dataframe should look like this:

```{r, echo = FALSE}
diamonds_means <- diamonds %>% 
  select(cut, x, y, z) %>% 
  group_by(cut) %>% 
  summarize(x_mean = mean(x),
            y_mean = mean(y),
            z_mean = mean(z))


diamonds_long <- diamonds_means %>% 
  pivot_longer(-cut, names_to = "coordinate", values_to = "mean_coord")

diamonds_wide <-  diamonds_long %>% 
    pivot_wider(names_from = cut, values_from = mean_coord)
  
diamonds_wide
```


```{r, eval = FALSE}
diamonds_means <- diamonds %>% 
  select(cut, x, y, z) %>% 
  group_by(___) %>% 
  summarize(x_mean = ____,
            y_mean = ____,
            z_mean = ____)


diamonds_reshaped <- diamonds_means %>% 
  pivot_ ______(____, names_ ___ = __________, values_ ___ = ___________) %>% 
  pivot_ ________(names_ ____ = ________, values_ ____ = __________)
  
diamonds_reshaped
```


### Handle dates with lubridate

Often enough, data are collected on the same subjects (in this case storm names) across time, this implied the use of dates. However, handling dates can become a bit of a headache if R does not recognize the format. 
The package lubridate helps specifically with this problem and allows to format dates in a way that can be unequivocally recognized by R.


Although lubridate is part of the tidyverse (and we do not need to install it separately), we need to call it separately because it does not get loaded when we load the tidyverse library.

```{r}
library(lubridate)
```

The best way to input dates (if you are the one doing the data input), is to create three separate columns for day, month, year.

Then we can use the function `ymd()` from the package lubridate in combination with the function `paste()` to create a date variable

```{r}
storms_with_date <- storms %>% 
  mutate(date = ymd(paste(year, month, day, sep = "-") )) 
```

if you query the new dataframe using the function `View()` you can see that the new variable date is specifically formatted as a Date data type.

```{r}
str(storms_with_date)
```


<!-- ### Working with factors -->
<!-- Dataframe often contain factors. Factors are categorical variables that are assigned different levels, but are read by R as numbers.  -->
<!-- The package forcats provide some useful functions that facilitate working with factors.  -->
<!-- Here we illustrate just a few of them -->


<!-- ### convert categorical variables into factors -->

<!-- ```{r} -->
<!-- storms_cat <- storms %>%  -->
<!--   mutate(category_factor = as_factor(category))  -->
<!-- ``` -->

<!-- ### Group levels of a factor together with fct_collapse -->

<!-- ```{r} -->
<!-- storms_cat <- storms_cat %>%  -->
<!--   mutate(broad_category = fct_collapse(category_factors,  -->
<!--                                        storm = c("-1","0"), -->
<!--                                        hurricane = c("1","2","3","4","5"))) -->
<!-- ``` -->

<!-- ### Change the name of the levels of a factor with fct_recode -->

<!-- ```{r} -->
<!-- storms_cat <- storms_cat %>%  -->
<!--   mutate(category_factor_long = fct_recode(broad_category, -->
<!--                                            "tropical storm" = "storm", -->
<!--                                            "hurricane" = "hurricane")) -->
<!-- ``` -->

